---
title: "Loss Reserving in Auto Liability: Chain Ladder & Bornhuetter-Ferguson"
author: "Christopher Flynn"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
editor_options: 
  markdown: 
    wrap: 72
---

```{r}

```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```

------------------------------------------------------------------------

## Introduction

When an auto liability insurance claim is reported, the insurer rarely
knows exactly how much it will ultimately cost. A fender-bender might
settle quickly; a serious injury claim could take years to resolve. This
creates a fundamental challenge for insurance companies: **how do you
set aside enough money today to pay claims that won't fully develop
until the future?**

The answer lies in **loss reserving** — a set of actuarial techniques
that use historical claims data to estimate how much money an insurer
will ultimately owe. This project walks through two widely used methods:

1.  **Chain Ladder (Development Method)** — projects future losses by
    assuming historical payment patterns repeat themselves.
2.  **Bornhuetter-Ferguson (BF)** — blends the Chain Ladder projection
    with an independent estimate of expected losses, giving more
    stability for immature accident years.

We'll build everything from scratch: generating synthetic claims data,
constructing a loss triangle, and applying both methods — all in
reproducible R code.

------------------------------------------------------------------------

## Glossary of Key Terms

Before diving into the data, here's a plain-language guide to the
terminology used throughout this analysis.

| Term | Definition |
|------------------------------------|------------------------------------|
| **Accident Year** | The calendar year in which a loss event (accident) occurred. Claims are grouped by when the accident happened, not when they were reported or paid. |
| **Development Lag** | How many years have passed since the accident year. Lag 1 = the year of the accident; Lag 2 = one year later; and so on. |
| **Paid Losses** | The actual dollars an insurer has paid out on claims so far — cash out the door. |
| **Reported (Incurred) Losses** | Paid losses *plus* case reserves — the insurer's best estimate of what each open claim will ultimately cost. |
| **Loss Triangle** | A matrix where rows are accident years and columns are development lags. Each cell shows cumulative losses as of a given lag. The upper-left portion is observed; the lower-right is unknown and must be estimated. |
| **Age-to-Age Factor (LDF)** | A multiplier that describes how much losses grow from one development lag to the next. Also called a Loss Development Factor. |
| **Tail Factor** | A factor applied beyond the last observed development period to capture any remaining loss development. |
| **Ultimate Losses** | The total amount an insurer expects to pay on a group of claims once all development is complete. |
| **IBNR** | Incurred But Not Reported — losses that have occurred but haven't yet been fully recognized in the insurer's books. IBNR = Ultimate − Reported to Date. |
| **A Priori Loss Ratio** | An independent, upfront estimate of losses as a percentage of premium. Used in the BF method. |
| **Expected Losses** | Premium × A Priori Loss Ratio. The baseline estimate of what losses *should* be before looking at actual data. |

------------------------------------------------------------------------

## Part 1: Generating Synthetic Claims Data

### Why Synthetic Data?

Real insurance claims data is confidential and proprietary. For this
portfolio demonstration, we'll simulate a realistic dataset for a
fictitious auto liability book of business. The data is designed to
mimic real-world patterns: losses develop slowly in early lags, then
taper off as claims close.

### The Simulation

We'll generate **individual claims** with the following fields:

-   `claim_id` — a unique identifier for each claim
-   `accident_year` — the year the accident occurred (2017–2024)
-   `development_lag` — years of development observed (1–8)
-   `paid_losses` — cumulative paid losses as of that lag
-   `reported_losses` — cumulative reported (incurred) losses as of that
    lag
-   `claim_count` — number of claims in that accident year / lag
    combination
-   `premium` — earned premium for that accident year (used in BF)

```{r generate-data}
library(tidyverse)
library(knitr)
library(kableExtra)
library(scales)

set.seed(42)  # Ensures reproducibility — anyone running this gets the same numbers

# ── Parameters ──────────────────────────────────────────────────────────────────
accident_years  <- 2017:2024
n_lags          <- 8
n_claims_range  <- c(180, 320)   # Claims per accident year

# Premium grows modestly each year (reflects a growing book of business)
premiums <- tibble(
  accident_year = accident_years,
  premium       = round(seq(8e6, 11.5e6, length.out = 8), -3)
)

# ── Development pattern ──────────────────────────────────────────────────────────
# These cumulative % factors describe how much of ultimate losses are paid by each lag.
# Auto liability tends to develop faster than, say, workers' comp.
cum_pct_paid <- c(0.38, 0.62, 0.76, 0.85, 0.91, 0.95, 0.98, 1.00)
cum_pct_rptd <- c(0.65, 0.82, 0.90, 0.95, 0.97, 0.98, 0.99, 1.00)

# ── Ultimate loss ratio assumptions ─────────────────────────────────────────────
# Each accident year has a slightly different underlying loss ratio, reflecting
# real-world variability in claim severity and frequency.
ult_loss_ratios <- c(0.68, 0.71, 0.74, 0.70, 0.73, 0.75, 0.72, 0.70)

# ── Build the claims dataset ─────────────────────────────────────────────────────
claims_list <- list()

for (i in seq_along(accident_years)) {
  ay      <- accident_years[i]
  prem    <- premiums$premium[i]
  ult_lr  <- ult_loss_ratios[i]
  ult_loss <- prem * ult_lr

  # The current calendar year is 2024, so accident year 2017 has 8 lags of
  # development, 2018 has 7, ..., 2024 has only 1.
  max_lag <- n_lags - (i - 1)   # most-recent AY has fewest lags observed

  n_claims <- sample(seq(n_claims_range[1], n_claims_range[2]), 1)

  for (lag in 1:max_lag) {
    # Add realistic noise around the expected development percentages
    noise_paid <- rnorm(1, 0, 0.015)
    noise_rptd <- rnorm(1, 0, 0.010)

    paid_pct <- min(max(cum_pct_paid[lag] + noise_paid, 0.01), 1.00)
    rptd_pct <- min(max(cum_pct_rptd[lag] + noise_rptd, paid_pct), 1.00)

    claims_list[[length(claims_list) + 1]] <- tibble(
      claim_id      = paste0("AY", ay, "-L", lag),
      accident_year = ay,
      development_lag = lag,
      claim_count   = round(n_claims * paid_pct),
      paid_losses   = round(ult_loss * paid_pct, 0),
      reported_losses = round(ult_loss * rptd_pct, 0),
      premium       = prem
    )
  }
}

claims_df <- bind_rows(claims_list)

# Preview the first several rows
claims_df %>%
  head(16) %>%
  mutate(across(c(paid_losses, reported_losses, premium), dollar)) %>%
  kable(caption = "Sample of Synthetic Auto Liability Claims Data",
        align   = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r export-csv}
# Export the full dataset as a CSV for sharing / reproducibility
write_csv(claims_df, "auto_liability_claims.csv")
cat("✓ Exported", nrow(claims_df), "rows to auto_liability_claims.csv\n")
```

A quick summary of what we just built:

```{r data-summary}
claims_df %>%
  group_by(accident_year) %>%
  summarise(
    lags_observed   = max(development_lag),
    total_claims    = max(claim_count),
    paid_at_latest  = dollar(max(paid_losses)),
    rptd_at_latest  = dollar(max(reported_losses)),
    premium         = dollar(first(premium))
  ) %>%
  kable(caption = "Summary by Accident Year (latest diagonal)",
        col.names = c("Accident Year", "Lags Observed", "# Claims",
                      "Paid to Date", "Reported to Date", "Premium"),
        align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

------------------------------------------------------------------------

## Part 2: Building the Loss Triangle

### What Is a Loss Triangle?

A **loss triangle** is the backbone of most reserving analyses. Imagine
laying out your claims data in a grid:

-   **Rows** = Accident Years (2017 through 2024)
-   **Columns** = Development Lags (Year 1 through Year 8)
-   **Each cell** = Cumulative losses as of that lag for that accident
    year

The result looks like a triangle because more-recent accident years have
fewer development periods observed — we can only see data through the
current calendar year. The lower-right portion of the grid is *blank* —
that's exactly what we need to estimate.

```{r build-triangle}
# ── Paid Loss Triangle ───────────────────────────────────────────────────────────
paid_triangle <- claims_df %>%
  select(accident_year, development_lag, paid_losses) %>%
  pivot_wider(
    names_from   = development_lag,
    values_from  = paid_losses,
    names_prefix = "Lag_"
  ) %>%
  arrange(accident_year) %>%
  column_to_rownames("accident_year")

# ── Reported Loss Triangle ───────────────────────────────────────────────────────
rptd_triangle <- claims_df %>%
  select(accident_year, development_lag, reported_losses) %>%
  pivot_wider(
    names_from   = development_lag,
    values_from  = reported_losses,
    names_prefix = "Lag_"
  ) %>%
  arrange(accident_year) %>%
  column_to_rownames("accident_year")

# Display the paid triangle with NA cells shown as "—"
paid_triangle %>%
  mutate(across(everything(), ~ ifelse(is.na(.), "—", dollar(.)))) %>%
  kable(caption = "Cumulative Paid Loss Triangle — Auto Liability (2017–2024)",
        align   = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  column_spec(1:8, width = "10%")
```

```{r display-rptd-triangle}
rptd_triangle %>%
  mutate(across(everything(), ~ ifelse(is.na(.), "—", dollar(.)))) %>%
  kable(caption = "Cumulative Reported Loss Triangle — Auto Liability (2017–2024)",
        align   = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE)
```

Notice the staircase pattern: accident year 2017 has all 8 lags filled
in, while accident year 2024 has only Lag 1 visible. The goal of our
reserving methods is to fill in the missing cells and sum up what we
expect each accident year to ultimately cost.

------------------------------------------------------------------------

## Part 3: Chain Ladder Method

### How Chain Ladder Works

The **Chain Ladder method** (also called the Development Method) is
built on one key assumption: *the way losses develop over time in the
past will continue in the future.*

The process has three steps:

1.  **Calculate Age-to-Age Factors (LDFs)** — for each pair of
    consecutive lags, how much did cumulative losses multiply from one
    lag to the next? Average these across all accident years to get a
    single factor per transition.
2.  **Apply the factors forward** — starting from the latest known
    diagonal, multiply each accident year's current losses by the
    appropriate LDFs to project out to ultimate.
3.  **Compute IBNR** — Ultimate minus what's already been reported = the
    reserve needed.

### Step 1: Calculate Age-to-Age Factors

```{r cl-ldfs}
# ── Volume-weighted age-to-age factors ──────────────────────────────────────────
# Volume-weighting sums numerators and denominators separately before dividing.
# This gives more weight to larger accident years and is the actuarial standard.

compute_ldfs <- function(triangle) {
  n_rows <- nrow(triangle)
  n_cols <- ncol(triangle)
  ldfs   <- numeric(n_cols - 1)

  for (j in 1:(n_cols - 1)) {
    # Only use rows where BOTH the current and next lag are observed
    numerator   <- sum(triangle[, j + 1], na.rm = TRUE)
    denominator <- sum(triangle[1:(n_rows - j), j], na.rm = TRUE)
    ldfs[j]     <- numerator / denominator
  }
  ldfs
}

paid_ldfs <- compute_ldfs(as.matrix(paid_triangle))
rptd_ldfs <- compute_ldfs(as.matrix(rptd_triangle))

# Add a tail factor of 1.000 (assuming development is complete at Lag 8)
tail_factor <- 1.000
paid_ldfs_full <- c(paid_ldfs, tail_factor)
rptd_ldfs_full <- c(rptd_ldfs, tail_factor)

ldf_table <- tibble(
  Transition  = c(paste0("Lag ", 1:7, " → ", 2:8), "Lag 8 → Ult"),
  `Paid LDF`  = round(paid_ldfs_full, 4),
  `Rptd LDF`  = round(rptd_ldfs_full, 4)
)

ldf_table %>%
  kable(caption = "Age-to-Age Loss Development Factors (LDFs)",
        align   = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

> **Reading the table:** A Paid LDF of, say, 1.63 from Lag 1 → Lag 2
> means that, on average, cumulative paid losses grew by 63% between the
> first and second year of development. Factors closer to 1.000 in later
> lags indicate that most losses have already been paid and little
> development remains.

### Step 2 & 3: Project to Ultimate and Calculate IBNR

```{r cl-ultimate}
# ── Cumulative-to-Ultimate (CDF) factors ────────────────────────────────────────
# The CDF at lag j = product of all LDFs from lag j onward.
# This tells us: "multiply current losses by this factor to get ultimate."

compute_cdfs <- function(ldfs_full) {
  n <- length(ldfs_full)
  cdfs <- numeric(n)
  cdfs[n] <- ldfs_full[n]
  for (j in (n - 1):1) {
    cdfs[j] <- ldfs_full[j] * cdfs[j + 1]
  }
  cdfs
}

paid_cdfs <- compute_cdfs(paid_ldfs_full)
rptd_cdfs <- compute_cdfs(rptd_ldfs_full)

# ── Extract the latest diagonal ──────────────────────────────────────────────────
# For each accident year, find its most recent observed value.
get_latest_diagonal <- function(triangle) {
  apply(triangle, 1, function(row) {
    vals <- row[!is.na(row)]
    if (length(vals) == 0) NA else tail(vals, 1)
  })
}

latest_paid <- get_latest_diagonal(as.matrix(paid_triangle))
latest_rptd <- get_latest_diagonal(as.matrix(rptd_triangle))

# The lag at which each accident year currently sits
current_lag <- sapply(as.data.frame(t(paid_triangle)), function(col) {
  sum(!is.na(col))
})
current_lag <- apply(as.matrix(paid_triangle), 1, function(row) sum(!is.na(row)))

# ── Compute ultimate and IBNR ────────────────────────────────────────────────────
cl_results <- tibble(
  accident_year       = accident_years,
  current_lag         = current_lag,
  paid_to_date        = latest_paid,
  reported_to_date    = latest_rptd,
  paid_cdf            = paid_cdfs[current_lag],
  rptd_cdf            = rptd_cdfs[current_lag],
  cl_ultimate_paid    = round(latest_paid * paid_cdfs[current_lag], 0),
  cl_ultimate_rptd    = round(latest_rptd * rptd_cdfs[current_lag], 0),
  ibnr_paid_basis     = round(latest_paid * paid_cdfs[current_lag] - latest_rptd, 0),
  ibnr_rptd_basis     = round(latest_rptd * rptd_cdfs[current_lag] - latest_rptd, 0)
)

cl_results %>%
  mutate(
    paid_to_date     = dollar(paid_to_date),
    reported_to_date = dollar(reported_to_date),
    paid_cdf         = round(paid_cdf, 4),
    rptd_cdf         = round(rptd_cdf, 4),
    cl_ultimate_rptd = dollar(cl_ultimate_rptd),
    ibnr_rptd_basis  = dollar(ibnr_rptd_basis)
  ) %>%
  select(accident_year, current_lag, reported_to_date, rptd_cdf,
         cl_ultimate_rptd, ibnr_rptd_basis) %>%
  kable(
    caption = "Chain Ladder Results — Reported Basis",
    col.names = c("Accident Year", "Current Lag", "Reported to Date",
                  "CDF to Ult", "CL Ultimate", "CL IBNR"),
    align = "c"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r cl-total}
cat("Chain Ladder Total IBNR (Reported Basis):",
    dollar(sum(cl_results$ibnr_rptd_basis)), "\n")
cat("Chain Ladder Total Ultimate (Reported Basis):",
    dollar(sum(cl_results$cl_ultimate_rptd)), "\n")
```

------------------------------------------------------------------------

## Part 4: Bornhuetter-Ferguson Method

### Why BF?

The Chain Ladder method is powerful, but it has a known weakness: **for
very recent accident years with little development, it leans heavily on
immature data.** If accident year 2024 has only one lag of history,
multiplying it by a large CDF amplifies any early-period noise.

The **Bornhuetter-Ferguson (BF) method** was introduced by Ronald
Bornhuetter and Ronald Ferguson in a landmark 1972 paper. Their insight:
blend two sources of information.

-   **What the data says** — actual losses reported to date (like Chain
    Ladder uses).
-   **What we expected** — a priori expected losses based on premium and
    an assumed loss ratio, set independently before looking at the data.

The BF ultimate estimate is:

> **BF Ultimate = Reported to Date + Expected Unreported Losses**

Where *Expected Unreported Losses* = Expected Ultimate × (1 − 1/CDF).

The term `(1 − 1/CDF)` is the **percent unreported** — how much of
ultimate losses we'd expect to *not yet be* in the data at the current
lag.

For mature accident years (CDF near 1.0), BF gives nearly the same
answer as Chain Ladder. For immature years, BF is more stable because it
leans on the a priori expectation rather than extrapolating from thin
data.

### A Priori Expected Losses

```{r bf-setup}
# ── A priori loss ratios ─────────────────────────────────────────────────────────
# In practice, these would come from pricing, industry benchmarks, or management
# judgment. Here we use slightly smoothed versions of our true simulation ratios.
apriori_lr <- c(0.70, 0.70, 0.72, 0.72, 0.73, 0.73, 0.72, 0.71)

bf_inputs <- cl_results %>%
  mutate(
    premium           = premiums$premium,
    apriori_lr        = apriori_lr,
    expected_ultimate = round(premium * apriori_lr, 0),
    pct_unreported    = round(1 - 1 / rptd_cdf, 4),   # Expected % still to develop
    expected_unreptd  = round(expected_ultimate * pct_unreported, 0)
  )

bf_inputs %>%
  mutate(
    premium           = dollar(premium),
    expected_ultimate = dollar(expected_ultimate),
    expected_unreptd  = dollar(expected_unreptd)
  ) %>%
  select(accident_year, premium, apriori_lr, expected_ultimate,
         pct_unreported, expected_unreptd) %>%
  kable(
    caption = "BF Method — A Priori Inputs",
    col.names = c("Accident Year", "Premium", "A Priori LR",
                  "Expected Ultimate", "% Unreported", "Expected Unreported"),
    align = "c"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### BF Ultimate and IBNR

```{r bf-ultimate}
bf_results <- bf_inputs %>%
  mutate(
    bf_ultimate = round(reported_to_date + expected_unreptd, 0),
    bf_ibnr     = round(bf_ultimate - reported_to_date, 0)
  )

bf_results %>%
  mutate(
    reported_to_date = dollar(reported_to_date),
    expected_unreptd = dollar(expected_unreptd),
    bf_ultimate      = dollar(bf_ultimate),
    bf_ibnr          = dollar(bf_ibnr)
  ) %>%
  select(accident_year, reported_to_date, expected_unreptd, bf_ultimate, bf_ibnr) %>%
  kable(
    caption = "Bornhuetter-Ferguson Results — Reported Basis",
    col.names = c("Accident Year", "Reported to Date", "Expected Unreported",
                  "BF Ultimate", "BF IBNR"),
    align = "c"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r bf-total}
cat("BF Total IBNR:", dollar(sum(bf_results$bf_ibnr)), "\n")
cat("BF Total Ultimate:", dollar(sum(bf_results$bf_ultimate)), "\n")
```

------------------------------------------------------------------------

## Part 5: Method Comparison & Visualization

### Side-by-Side: Chain Ladder vs. BF

```{r comparison-table}
comparison <- tibble(
  accident_year    = accident_years,
  reported_to_date = cl_results$reported_to_date,
  cl_ultimate      = cl_results$cl_ultimate_rptd,
  cl_ibnr          = cl_results$ibnr_rptd_basis,
  bf_ultimate      = bf_results$bf_ultimate,
  bf_ibnr          = bf_results$bf_ibnr,
  diff_ibnr        = bf_results$bf_ibnr - cl_results$ibnr_rptd_basis
)

comparison %>%
  mutate(across(c(reported_to_date, cl_ultimate, cl_ibnr,
                  bf_ultimate, bf_ibnr, diff_ibnr), dollar)) %>%
  kable(
    caption = "Chain Ladder vs. Bornhuetter-Ferguson — Full Comparison",
    col.names = c("Accident Year", "Reported to Date",
                  "CL Ultimate", "CL IBNR",
                  "BF Ultimate", "BF IBNR",
                  "BF − CL IBNR"),
    align = "c"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(which(abs(bf_results$bf_ibnr - cl_results$ibnr_rptd_basis) ==
                   max(abs(bf_results$bf_ibnr - cl_results$ibnr_rptd_basis))),
           bold = TRUE, background = "#fff3cd")
```

### Ultimate Loss Estimates by Accident Year

```{r plot-ultimates, fig.cap="Chain Ladder vs. BF Ultimate Estimates by Accident Year"}
plot_data <- comparison %>%
  select(accident_year, cl_ultimate, bf_ultimate) %>%
  pivot_longer(cols = c(cl_ultimate, bf_ultimate),
               names_to = "method", values_to = "ultimate") %>%
  mutate(method = recode(method,
                         "cl_ultimate" = "Chain Ladder",
                         "bf_ultimate" = "Bornhuetter-Ferguson"))

ggplot(plot_data, aes(x = factor(accident_year), y = ultimate, fill = method)) +
  geom_col(position = "dodge", width = 0.65, alpha = 0.88) +
  geom_text(aes(label = paste0("$", round(ultimate / 1e6, 2), "M")),
            position = position_dodge(width = 0.65),
            vjust = -0.4, size = 3.2, fontface = "bold") +
  scale_y_continuous(labels = label_dollar(scale = 1e-6, suffix = "M"),
                     expand = expansion(mult = c(0, 0.12))) +
  scale_fill_manual(values = c("Chain Ladder" = "#2c7bb6", "Bornhuetter-Ferguson" = "#d7191c")) +
  labs(
    title    = "Projected Ultimate Losses by Accident Year",
    subtitle = "Auto Liability | Chain Ladder vs. Bornhuetter-Ferguson",
    x        = "Accident Year",
    y        = "Ultimate Losses",
    fill     = "Method",
    caption  = "Note: Based on synthetic data generated for illustration purposes."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "grey40"),
    legend.position = "top",
    panel.grid.major.x = element_blank()
  )
```

### IBNR by Accident Year

```{r plot-ibnr, fig.cap="IBNR Estimates: Chain Ladder vs. BF"}
ibnr_data <- comparison %>%
  select(accident_year, cl_ibnr, bf_ibnr) %>%
  pivot_longer(cols = c(cl_ibnr, bf_ibnr),
               names_to = "method", values_to = "ibnr") %>%
  mutate(method = recode(method,
                         "cl_ibnr" = "Chain Ladder",
                         "bf_ibnr" = "Bornhuetter-Ferguson"))

ggplot(ibnr_data, aes(x = factor(accident_year), y = ibnr, fill = method)) +
  geom_col(position = "dodge", width = 0.65, alpha = 0.88) +
  scale_y_continuous(labels = label_dollar(scale = 1e-6, suffix = "M"),
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_manual(values = c("Chain Ladder" = "#2c7bb6", "Bornhuetter-Ferguson" = "#d7191c")) +
  labs(
    title    = "IBNR Reserves by Accident Year",
    subtitle = "Difference is largest for the most recent (immature) accident years",
    x        = "Accident Year",
    y        = "IBNR",
    fill     = "Method",
    caption  = "IBNR = Estimated Ultimate − Reported to Date"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "grey40"),
    legend.position = "top",
    panel.grid.major.x = element_blank()
  )
```

### Development Pattern Visualization

```{r plot-development, fig.cap="Loss Development Pattern by Lag"}
dev_data <- claims_df %>%
  group_by(accident_year, development_lag) %>%
  summarise(reported_losses = max(reported_losses), .groups = "drop") %>%
  group_by(accident_year) %>%
  mutate(pct_of_latest = reported_losses / max(reported_losses)) %>%
  ungroup()

ggplot(dev_data, aes(x = development_lag, y = reported_losses / 1e6,
                     color = factor(accident_year), group = accident_year)) +
  geom_line(linewidth = 1.1, alpha = 0.8) +
  geom_point(size = 2.5) +
  scale_color_brewer(palette = "Dark2") +
  scale_x_continuous(breaks = 1:8) +
  scale_y_continuous(labels = label_dollar(suffix = "M")) +
  labs(
    title  = "Cumulative Reported Loss Development by Accident Year",
    x      = "Development Lag (Years)",
    y      = "Cumulative Reported Losses ($M)",
    color  = "Accident Year"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", size = 15),
    legend.position = "right"
  )
```

------------------------------------------------------------------------

## Part 6: Summary & Conclusions

```{r final-summary}
summary_tbl <- tibble(
  Metric              = c("Total Reported to Date", "Chain Ladder Ultimate",
                           "Chain Ladder IBNR", "BF Ultimate", "BF IBNR"),
  Value               = c(
    dollar(sum(cl_results$reported_to_date)),
    dollar(sum(cl_results$cl_ultimate_rptd)),
    dollar(sum(cl_results$ibnr_rptd_basis)),
    dollar(sum(bf_results$bf_ultimate)),
    dollar(sum(bf_results$bf_ibnr))
  )
)

summary_tbl %>%
  kable(caption = "Portfolio Summary — All Accident Years Combined", align = "lr") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(c(3, 5), bold = TRUE, background = "#e8f4f8")
```

### Key Takeaways

**Chain Ladder** is intuitive and data-driven, but it amplifies noise in
immature accident years. Notice how the Chain Ladder IBNR for 2023 and
2024 is more volatile — it's extrapolating aggressively from just one or
two lags of data.

**Bornhuetter-Ferguson** tempers that volatility by anchoring the
immature years to an a priori expectation. For 2024 (only one lag
visible), BF is largely driven by expected losses — which is the
appropriate behavior. As more data matures, BF naturally converges
toward the Chain Ladder result.

In practice, actuaries use **both methods** (and often others, like the
Benktander method or Cape Cod) and apply professional judgment to select
or blend estimates. No single method is always right — the goal is a
reasonable, defensible reserve that ensures the insurer can pay its
obligations to policyholders.

------------------------------------------------------------------------

## Session Info

```{r session-info}
sessionInfo()
```

------------------------------------------------------------------------

*This analysis uses fully synthetic data and is intended for educational
and portfolio demonstration purposes only. It does not represent the
financials of any real insurance company.*
